A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-29 paper/writeup.tex
Issues 10 and 14

There has been substantial recent progress
in the problem of ``ARG inference'',
which seeks to infer (portions of) the ancestral recombination graph (or, ARG)
that describes how a set of genotypes samples are related to each other
at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
One way of viewing the ARG is as a sequence of marginal trees,
i.e., the genealogical trees that describe how each portion of the genome
was inherited by the focal genomes.
This is reflected in methodology of some ARG inference methods
and in metrics used to assess inference accuracy,
as well as in basic terminology.
For instance, 
the ``succinct tree sequence'',
introduced by \citet{kelleher2016efficient},
is a common format for describing these inferred ARGs,
and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
\tskit{} \citep{tskit,ralph2020efficiently}.

However, the ARG is emphatically not merely a sequence of trees:
viewed another way, it describes inheritance relationships between ancestral haplotypes.
These two points of view are related because a single haplotype
may extend over many marginal trees;
in other words, the internal nodes in the trees are labeled, and many of these labels
are shared between adjacent trees.

@@

Haplotype identity has been largely overlooked in the literature on ARG inference --
most methods that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of marginal trees,
not on how ancestral haplotypes span across these trees.
For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
compared true and inferred ARGs
using average Robinson-Foulds and Kendall-Colijn distances between trees
across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure --
two ARGs could be identical by either measure
but imply completely different patterns of haplotype sharing and inheritance.
% (as defined in the context of the ARG, \citep{XXX}).
There has been some recent attention to haplotypes:
\citet{ignatieva2024length} compared distributions of haplotype spans in true and inferred ARGs,
as well as more sophisticated summaries of edges.
Also, \citet{deng2021distribution} evaluated agreement of distributions of
distances along the genome between tree topology changes, and
\citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
that is the total variation distance between the induced distribution on genotypes;
neither of which measure the sharing of haplotypes between adjacent trees.
However, the additional information provided by haplotype structure can be important:
for instance, haplotypes that extend over many marginal genealogies
``tie together'' those genealogies, allowing estimates of times of particular ancestors
to be informed by larger portions of the genome on which there are many genealogies.

In this paper, we study various aspects of haplotype identity in ARGs.
First, we describe a deterministic algorithm that
extends the genomic region spanned by ancestral haplotypes
using the principle that intermediate nodes in inheritance paths
should remain unchanged when possible.
These extended portions of ancestral haplotypes manifest as unary nodes in the marginal trees.
To quantify how accurate the new information is,
we define and describe how to compute new measures of (dis-)agreement between ARGs
that are motivated by the Robinson-Foulds distance between trees
but account for haplotype identity.
These measures show that the vast majority of these extended haplotypes are correct if the trees are correct,
and that substantial information about haplotypes is contained in these nodes
in inferred trees as well.

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=0.9\textwidth]{conceptual_figure.pdf}
    \end{center}
    \caption{
        A simple example showing the basic idea
        (described in more detail in the text):
        \textbf{(A)} a small portion of a ARG without unary nodes;
        \textbf{(B)} the implied inheritance pattern of the two portions of the haplotype carried by ancestral node 4,
        labeled $L$ and $R$;
        \textbf{(C)} marginal trees with a unary node added,
        which produces \textbf{(D)} a more parsimonious haplotype inheritance pattern
        (that also includes fewer edges).
        \label{fig:conceptual}
    }
\end{figure}

@@

% Description of algorithm
Given a tree sequence, our goal is to
identify areas of implied inheritance of haplotypes.
Generalizing from \figref{fig:extending_diagram},
we do this by identifying paths of inheritance that are shared across a sequence of marginal trees
but for which some of the intermediate nodes are missing.
Concretely, suppose that 
if in tree $T_k$ there is a chain of inheritance
$p \to u_1 \to \cdots \to u_m \to c$
(where $a \to b$ denotes a parent-child relationship)
and in tree $T_{k+1}$ there is a chain of inheritance
$p \to v_1 \to \cdots \to v_n \to c$,
where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
Of course, this does not always make sense --
for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
or if $t_{u_i} = t_{v_j}$, for some $i$ and $j$.
So, we restrict our attention to pairs of such paths in adjacent trees
for which
$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
$v_j \notin T_k$ for all $1 \le j \le n$,
and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
Call a pair of such paths \emph{mergeable}.
So, the goal of our algorithm is to iterate over trees,
identify mergeable pairs of paths,
and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)

@@

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Given a node $c$, trees $T_O$ and $T_I$,
        and sub-forests $F_O$ and $F_I$
        such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
        check to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
        If the paths are mergeable then this returns the number of new edges
        that would be added by extending the path from $T_O$ to $T_I$;
        otherwise, this returns $\infty$.
        Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
        in the set of edges to be removed and added, respectively
        (i.e., in $F_O$ and $F_I$).
        The variable $m_e$ will record the number of new edges to be added,
        and $m$ will record the number of extended haplotypes.
    }
	\label{alg:mergeable}
	\SetKwFunction{mergeable}{Mergeable}
	\DontPrintSemicolon
    \Fn{\mergeable($c$, $T_O$, $T_I$)}{
        \text{Let } $p_i = P_I[c]$, 
        $t_i = t[p_i]$,
        $p_o = P_O[c]$,
        $t_o = t[p_o]$,
        \text{ and }$m_e = m = 0$.\;
        \While{\tn{True}}{
            \text{Set } $y_i = (p_i \neq NULL)
                \& (p_i \notin T_O)
                \& (t_i < t_o)$ \;
            \text{and } $y_o = (p_o \neq NULL)
                \& (p_o \notin T_I)
                \& (t_o < t_i)$ \;
            \If{not ($y_i$ or $y_o$)}{ \Break\; }
            \eIf{$y_i$}{
                \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
                    \text{Set } $m_e = m_e + 1$.\;
                }
                \text{Set } $c = p_i$,
                $p_i = P_I[p_i]$, \text{ and}\;
                $t_i = \text{ if }(p_i = NULL)\text{ then }\infty\text{ else }t[p_i]$ .\; 
            }{
                \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
                    \text{Set } $m_e = m_e + 1$.\;
                }
                \text{Set }$c = p_o$,
                $p_o = P_O[p_o]$, \;
                $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
                \text{and } $m = m + 1$ .\;
            }
        }
        \If{$m = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
            \text{Set } $m_e = \infty$. \;
        }
        \KwRet{$m_e$}\;
    }
\end{algorithm2e}

@@

Concretely, our method can be applied to any two tree sequences,
as illustrated in \figref{fig:conceptual_discrepancy}.
However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
	\mathcal{M}(n1, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
to be the indices of all trees where $n_1$ and $n_2$ are ancestral to the same sample set in both ARGs.
Further, we define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k\in \mathcal{M}(n_1, n_2)} (a_k - a_{k-1}) ,  
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there some nodes in $\T_2$ may not be mapped to by any nodes in $\T_1$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
its' \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is not unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$,
and let $s(\T,n)$ denote
the total span that node $n$ is present in the marginal trees,
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
\end{align*}
where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise).
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
    = 
    \Vert \T_1\Vert -\similarity(\T_1,\T_2),
\end{align*}
which is the total span for all nodes in $\T_1$
over which their descendant samples do \emph{not} match those of their best match in $\T_2$.

Other authors \citep[e.g.,][]{kelleher2019inferring} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    \mathcal{M}'(\T_1, n; \T_2) = \left\{ k : \exists n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
the indices of all trees on which there is \emph{some} node in $\T_2$
whose set of descendant samples matches those of $n$, and
\begin{align*}
    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)}^N (a_k - a_{k-1})
\end{align*}
the total span over which $n$ finds a match.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree.
The other differences are that \emph{average} Robinson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
A similar metric on tree sequences could be defined
using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.

@@

To do this, we simulate ARGs containing full haplotypes using \msprime{}
\citep{kelleher2016efficient,baumdicker2021efficient},
with the \texttt{coalescing\_segments\_only} option set to False.
Although \msprime{} simulates many events that do not create a coalescence in some marginal tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in the ARG
output by \msprime{} are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the marginal trees.
Setting \texttt{coalescing\_segments\_only} to False includes just this information:
any ancestral segments for which these coalescent nodes
are ancestral to any samples -- so, the unary portions of their spans as well.
However, this includes more information than we want:
we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{wong2024general}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, after simulation, we first remove these isolated, non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.

@@

Our next task is to confirm that the haplotypes extended by \algorithmref{alg:extend}
are indeed correct -- i.e., that in addition to compression, we are also gaining information.
To do this, 
we simulate ARGs containing full haplotypes using \msprime{},
apply the simplification algorithm to reduce these so that there are no unary nodes
(i.e., any node present in a marginal tree is a coalescent node or a sample),
and then apply \algorithmref{alg:extend} to the result
(see Methods for more detail).
The method can potentially extend the spans of each node
(additional span over which the node will be unary);
and we can quantify how much of these extended spans were in the original ARG
(and thus correctly extended).

@@

\begin{figure}[!hbt]
	\begin{center}
        \includegraphics[width=\textwidth]{figure5.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
        and sequence lengths, with 1,000 diploid samples (bottom).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified}, removing the unary spans,
        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
        then each of these had its haplotypes \emph{extended}.
        The inferred, then simplified, ARG is also shown.
        \textbf{Top row:} ARF, the dissimilarity to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (\eqnref{eqn:dissimilarity}).
        \textbf{Bottom row:} TPR, agreement to the true ARG
        as proportion of the true ARG that is represented.
        }     
    \label{fig:dissimilarity}
\end{figure}

>>>>

There has been substantial recent progress
in the problem of ``ARG inference'',
which seeks to infer (portions of) the ancestral recombination graph (or, ARG)
that describes how a set of genotypes samples are related to each other
at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
One way of viewing the ARG is as a sequence of local trees,
i.e., the genealogical trees that describe how each portion of the genome
was inherited by the focal genomes.
This is reflected in methodology of some ARG inference methods
and in metrics used to assess inference accuracy,
as well as in basic terminology.
For instance, 
the ``succinct tree sequence'',
introduced by \citet{kelleher2016efficient},
is a common format for describing these inferred ARGs,
and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
\tskit{} \citep{tskit,ralph2020efficiently}.

However, the ARG is emphatically not merely a sequence of trees:
viewed another way, it describes inheritance relationships between ancestral haplotypes.
These two points of view are related because a single haplotype
may extend over many local trees;
in other words, the internal nodes in the trees are labeled, and many of these labels
are shared between adjacent trees.

@@

Haplotype identity has been largely overlooked in the literature on ARG inference --
most methods that have been used so far to measure accuracy of inferred ARGs
depend only on the sequence of local trees,
not on how ancestral haplotypes span across these trees.
For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
compared true and inferred ARGs
using average Robinson-Foulds and Kendall-Colijn distances between trees
across a regular sequence of genomic positions,
using sampled genotypes as labels,
while \citet{brandt2022evaluation} compared times to most recent common ancestor
between pairs of sampled genomes.
Neither is affected by shared haplotype structure --
two ARGs could be identical by either measure
but imply completely different patterns of haplotype sharing and inheritance.
% (as defined in the context of the ARG, \citep{XXX}).
There has been some recent attention to haplotypes:
\citet{ignatieva2024length} compared distributions of haplotype spans in true and inferred ARGs,
as well as more sophisticated summaries of edges.
Also, \citet{deng2021distribution} evaluated agreement of distributions of
distances along the genome between tree topology changes, and
\citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
that is the total variation distance between the induced distribution on genotypes;
neither of which measure the sharing of haplotypes between adjacent trees.
However, the additional information provided by haplotype structure can be important:
for instance, haplotypes that extend over many marginal genealogies
``tie together'' those genealogies, allowing estimates of times of particular ancestors
to be informed by larger portions of the genome on which there are many genealogies.

In this paper, we study various aspects of haplotype identity in ARGs.
First, we describe a deterministic algorithm that
extends the genomic region spanned by ancestral haplotypes
using the principle that intermediate nodes in inheritance paths
should remain unchanged when possible.
These extended portions of ancestral haplotypes manifest as unary nodes in the local trees.
To quantify how accurate the new information is,
we define and describe how to compute new measures of (dis-)agreement between ARGs
that are motivated by the Robinson-Foulds distance between trees
but account for haplotype identity.
These measures show that the vast majority of these extended haplotypes are correct if the trees are correct,
and that substantial information about haplotypes is contained in these nodes
in inferred trees as well.

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=0.9\textwidth]{conceptual_figure.pdf}
    \end{center}
    \caption{
        A simple example showing the basic idea
        (described in more detail in the text):
        \textbf{(A)} a small portion of a ARG without unary nodes;
        \textbf{(B)} the implied inheritance pattern of the two portions of the haplotype carried by ancestral node 4,
        labeled $L$ and $R$;
        \textbf{(C)} local trees with a unary node added,
        which produces \textbf{(D)} a more parsimonious haplotype inheritance pattern
        (that also includes fewer edges).
        \label{fig:conceptual}
    }
\end{figure}

@@

% Description of algorithm
Given a tree sequence, our goal is to
identify areas of implied inheritance of haplotypes.
Generalizing from \figref{fig:extending_diagram},
we do this by identifying paths of inheritance that are shared across a sequence of local trees
but for which some of the intermediate nodes are missing.
Concretely, suppose that 
if in tree $T_k$ there is a chain of inheritance
$p \to u_1 \to \cdots \to u_m \to c$
(where $a \to b$ denotes a parent-child relationship)
and in tree $T_{k+1}$ there is a chain of inheritance
$p \to v_1 \to \cdots \to v_n \to c$,
where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
Of course, this does not always make sense --
for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
or if $t_{u_i} = t_{v_j}$, for some $i$ and $j$.
So, we restrict our attention to pairs of such paths in adjacent trees
for which
$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
$v_j \notin T_k$ for all $1 \le j \le n$,
and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
Call a pair of such paths \emph{mergeable}.
So, the goal of our algorithm is to iterate over trees,
identify mergeable pairs of paths,
and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)

@@

\begin{algorithm2e}[!ht]
	\SetStartEndCondition{ }{}{}%
	\SetKwProg{Fn}{def}{\string:}{}
	\SetKwFunction{Range}{range}%%
	\SetKw{KwTo}{in}
	\SetKwFor{For}{for}{\string:}{}%
	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
	\SetKwFor{While}{while}{:}{fintq}%
	\SetKw{Break}{break}
	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
    \caption{
        Given a node $c$, trees $T_O$ and $T_I$,
        and sub-forests $F_O$ and $F_I$
        such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
        check to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
        If the paths are mergeable then this returns the number of new edges
        that would be added by extending the path from $T_O$ to $T_I$;
        otherwise, this returns $\infty$.
        Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
        in the set of edges to be removed and added, respectively
        (i.e., in $F_O$ and $F_I$).
        The variable $m_e$ will record the number of new edges to be added,
        and $m$ will record the number of extended haplotypes.
    }
	\label{alg:mergeable}
	\SetKwFunction{mergeable}{Mergeable}
	\DontPrintSemicolon
    \Fn{\mergeable($c$, $T_O$, $T_I$)}{
        \text{Let } $p_i = P_I[c]$, 
        $t_i = t[p_i]$,
        $p_o = P_O[c]$,
        $t_o = t[p_o]$,
        \text{ and }$m_e = m = 0$.\;
        \While{\tn{True}}{
            \text{Set } $y_i = (p_i \neq NULL)
                \ \&\ (p_i \notin T_O)
                \ \&\ (t_i < t_o)$ \;
            \text{and } $y_o = (p_o \neq NULL)
                \ \&\ (p_o \notin T_I)
                \ \&\ (t_o < t_i)$ \;
            \If{not ($y_i$ or $y_o$)}{ \Break\; }
            \eIf{$y_i$}{
                \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
                    \text{Set } $m_e = m_e + 1$.\;
                }
                \text{Set } $c = p_i$,
                $p_i = P_I[p_i]$, \text{ and}\;
                $t_i = \text{ if }(p_i = NULL)\text{ then }\infty\text{ else }t[p_i]$ .\; 
            }{
                \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
                    \text{Set } $m_e = m_e + 1$.\;
                }
                \text{Set }$c = p_o$,
                $p_o = P_O[p_o]$, \;
                $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
                \text{and } $m = m + 1$ .\;
            }
        }
        \If{$m = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
            \text{Set } $m_e = \infty$. \;
        }
        \KwRet{$m_e$}\;
    }
\end{algorithm2e}

@@

Concretely, our method can be applied to any two tree sequences,
as illustrated in \figref{fig:conceptual_discrepancy}.
However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
	\mathcal{M}(n1, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
to be the indices of all trees where $n_1$ and $n_2$ are ancestral to the same sample set in both ARGs.
Further, we define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k\in \mathcal{M}(n_1, n_2)} (a_k - a_{k-1}) ,  
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there some nodes in $\T_2$ may not be mapped to by any nodes in $\T_1$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
its' \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is not unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$,
and let $s(\T,n)$ denote
the total span that node $n$ is present in the local trees,
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
\end{align*}
where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise).
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
    = 
    \Vert \T_1\Vert -\similarity(\T_1,\T_2),
\end{align*}
which is the total span for all nodes in $\T_1$
over which their descendant samples do \emph{not} match those of their best match in $\T_2$.

Other authors \citep[e.g.,][]{kelleher2019inferring} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    \mathcal{M}'(\T_1, n; \T_2) = \left\{ k : \exists\ n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
the indices of all trees on which there is \emph{some} node in $\T_2$
whose set of descendant samples matches those of $n$, and
\begin{align*}
    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)} (a_k - a_{k-1})
\end{align*}
the total span over which $n$ finds a match.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree.
The other differences are that \emph{average} Robinson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
A similar metric on tree sequences could be defined
using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.

@@

Our new method \textit{extend haplotypes} is applicable to any ARG.
However its accuracy depends on the overall structure of the ARG it is applied too. 
Additionally, as \textit{ARG Robinson-Foulds}
is given as a proportion of node spans
it's value is also dependent on the pair of input ARGs. 
Thus to understand how well our methods can infer ancestral haplotypes we 
work with simulated tree sequences with a large variety in parameter choice.
To do this, we simulate ARGs containing full haplotypes using \msprime{}
\citep{kelleher2016efficient,baumdicker2021efficient},
with the \texttt{coalescing\_segments\_only} option set to False.
Although \msprime{} simulates many events that do not create a coalescence in some local tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in the ARG
output by \msprime{} are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the local trees.
Setting \texttt{coalescing\_segments\_only} to False includes just this information:
any ancestral segments for which these coalescent nodes
are ancestral to any samples -- so, the unary portions of their spans as well.
However, this includes more information than we want:
we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{wong2024general}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, after simulation, we first remove these isolated, non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.

@@

Our next task is to confirm that the haplotypes extended by \algorithmref{alg:extend}
are indeed correct -- i.e., that in addition to compression, we are also gaining information.
To do this, 
we simulate ARGs containing full haplotypes using \msprime{},
apply the simplification algorithm to reduce these so that there are no unary nodes
(i.e., any node present in a local tree is a coalescent node or a sample),
and then apply \algorithmref{alg:extend} to the result
(see Methods for more detail).
The method can potentially extend the spans of each node
(additional span over which the node will be unary);
and we can quantify how much of these extended spans were in the original ARG
(and thus correctly extended).

@@

\begin{figure}[!hbt]
	\begin{center}
        \includegraphics[width=\textwidth]{figure5.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (right)
        and sequence lengths, with 1,000 diploid samples (left).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified}, removing the unary spans,
        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
        then each of these had its haplotypes \emph{extended}.
        The inferred, then simplified, ARG is also shown.
        \textbf{Top row:} ARF, the dissimilarity to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (\eqnref{eqn:dissimilarity}).
        \textbf{Bottom row:} TPR, agreement to the true ARG
        as proportion of the true ARG that is represented.
        }     
    \label{fig:dissimilarity}
\end{figure}
